SDGs

### 2022-12-02 inequality-SDGs.Rmd
progress on inequality SDG analysis at national scale + matching problem

@@

```
### Correlations

@@

```{r}
dat |>
    filter(indicator_name %in% vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) |> 
    GGally::ggpairs(
        columns = 4:12 ,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 6)
```
Interpolate missing values

```{r}
df_dat <- dat |>
    filter(indicator_name %in% vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) %>% 
    group_by(country_code) %>% 
    mutate(across(.cols = vars, .fns = na_interpolation, option = "spline"))

df_dat
```

@@

Why are there more than one obs per country per year per indicator???
```{r}
sdg |> 
    filter(`ISO-alpha3 Code` %in% countries_list) |> 
    group_by(geo_area_code, indicator, time_period) |> 
    summarize(n = n()) |> 
    filter(n>19) |>  # 19 is the max number of years
    arrange(desc(n))

@@

```{r}
sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    group_by(indicator, indicator_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |>
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
     mutate(indicator_name = str_trunc(indicator_name, width = 30, side = "right")) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    labs(title = "SDG15: Life on land") +
    theme_light(base_size = 8)

sdg|> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    pull(indicator_name) |> unique()

@@

@@

Select the variables for which you have time series. Repeat the trajectories analysis. 
Do static analysis with all SDGs

```{r}
ctr <- sdg |> pull(geo_area_name) |> unique() |> length()
sdg <- sdg |> select(-c(age:type_of_speed))

sdg <- sdg |> 
    mutate(value = as.numeric(value)) |> 
    rename(year = time_period)
yrs <- sdg$year |> range() |> diff()
sdg <- sdg |> unique()

dat2 <- sdg |> 
    filter(goal == 10 | goal == 14 | goal == 15 ) |> 
    group_by(indicator, geo_area_code, year, geo_area_name, goal, target, series_code, series_description) |> 
    summarize(value = mean(value))

## If I add series description it duplicate entries, possibly due to misspellings
#test
dat2 |>
    unique() |>
    add_count() |>
    filter(n>1)

dat2 |> 
    add_count() |> 
    ungroup() |> 
    group_by(geo_area_name,  series_code) |> 
    summarize(yrs_cov = sum(n)) |> 
    filter(yrs_cov > 20)

sdg |> filter(series_code == "DC_ODA_BDVDL")
```

>>>>

## World Bank SDGs

@@

q <- dat |>
    filter(indicator_name %in%  vars) |>
    select(-x35) |>
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |>
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |>
    mutate(nas = is.na(value)) |>
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    group_by(country_code, indicator_name) |>
    summarise(missing = sum(nas) / diff(range(year))) |>
    #filter(country_code %in% complete_countries) |> # this line only if next cell already run
    # ungroup() |> group_by(country_code) |>
    # summarize(mean_missing = mean(missing)) |>
    #filter(mean_missing < 0.3) |>
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90))

q

```{r}
complete_countries <-  dat |>
    filter(indicator_name %in%  vars) |>
    select(-x35) |>
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |>
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |>
    mutate(nas = is.na(value)) |>
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    group_by(country_code, indicator_name) |>
    summarise(missing = sum(nas) / diff(range(year))) |> 
    # filter out any country with missing values higher than 70%
    filter(!any(missing > 0.3)) |> 
    pull(country_code) |> unique()

complete_countries
```

Using `tsibble`:

```{r}

dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x")) |>  ## Formating for year-month-day
    #mutate(year = lubridate::parse_date_time(year, orders = "Y", select_formats = "%Y"))  |> 
    #mutate(year = strptime(year, format = "%Y")) |>
    mutate(year = lubridate::make_date(year = year)) |> 
    as_tsibble(key = c(country_code, indicator_name), index=year, regular = TRUE) 
    # count_gaps()

### Correlations

@@

```{r}
dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) |> 
    GGally::ggpairs(
        columns = 4:12 ,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 6)
```
Interpolate missing values

```{r}
df_dat <- dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) %>% 
    group_by(country_code) %>% 
    mutate(across(.cols = vars, .fns = na_interpolation, option = "spline"))

df_dat |> ungroup() |> skimr::skim() # no missing values
```

@@

Why are there more than one obs per country per year per indicator???
```{r}
sdg |> 
    filter(`ISO-alpha3 Code` %in% countries_list) |> 
    group_by(geo_area_code, indicator, time_period) |> 
    summarize(n = n()) |> 
    filter(n>19) |>  # 19 is the max number of years
    arrange(desc(n)) #|> pull(geo_area_code) |> unique() # 191 countries (all of them)

@@

```{r}
sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    group_by(indicator, indicator_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |>
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator, indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    mutate(indicator_name = str_trunc(indicator_name, width = 30, side = "right")) |>
    ggplot(aes(year, indicator)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    labs(title = "SDG15: Life on land") +
    theme_light(base_size = 8)

## Very long names:
sdg|> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    select(indicator, indicator_name) |> 
    unique()

@@

Usable variables from UN Dataset:

```{r}
p2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(goal %in% c(10, 14, 15)) |> 
    group_by(series_code, year, iso3) |>
    # this gets rid of multiple values per year
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    ungroup() |> group_by(series_code, iso3) |> 
   # group_by(indicator, indicator_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year)) ) |>  ## dividing by number of years (per country)
    #pull(missing) |> range() # should be between 0-1
    # summarize(mean_missing = mean(missing)) |> 
    filter(missing < 0.3) |> 
    ungroup() |> 
    pivot_wider(names_from = series_code, values_from = missing) |> 
    pivot_longer(cols = AG_LND_FRSTCERT:last_col(), names_to = "series_code",
                 values_to = "missing") |> 
    group_by(series_code) |> 
    #filter(!any(is.na(missing))) |> 
    # ggplot(aes(iso3, series_code)) +
    # geom_tile(aes(fill = missing)) +
    # theme_light()
    summarize(missing_countries = sum(is.na(missing))) |> 
    filter(missing_countries < 45) |> 
    ggplot(aes(missing_countries, series_code))+
    geom_point()

p2 + theme_light(base_size = 12)

vars2 <- p2$data$series_code
```

```{r}
q2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(series_code %in% vars2) |> 
    ## this line is to get rid of multiple values per year
    group_by(series_code, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> #group_by(indicator_name, year, iso3) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(series_code, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year))) |>  ## dividing by number of years (per country)
    # summarize(mean_missing = mean(missing)) |> 
    ungroup() |> 
    filter(missing < 0.3) |> 
    ## add missing values again for combination of indicator country:
    ungroup() |> group_by(series_code) |>
    pivot_wider(names_from = iso3, values_from = missing) |> 
    pivot_longer(cols = AFG:last_col(), names_to = "iso3", values_to = "missing") |>
    ## completely empty time series
    #summarize(absent_countries = sum(is.na(missing))) |> 
    ungroup() |> group_by(iso3 ) |> 
    #filter(sum(is.na(missing)) < 45) |> # remove absent_countries > 45
    filter(!any(is.na(missing))) |> 
    ggplot(aes(iso3, series_code)) +
    #geom_point()
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90))

q2

complete_countries_sdg <- q2$data$iso3 |> unique()
```

Reduce the dataset to variables and countries with almost complete data. Then impute missing values.

```{r}
df_un_reduced <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries_sdg) |> # use only countries
    filter(series_code %in% vars2) |> 
    group_by(series_code, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    group_by(iso3, series_code) |> 
    mutate(value = na_interpolation(value, option = "spline"))

df_un_reduced |> ungroup() |> skimr::skim() # zero missing values
```

Now there is two working datasets cleaned and complete: `df_dat` is the data from World Bank, and `df_un_reduced` is the data from UN SDGs.

@@

1. Select the variables for which you have time series. Repeat the trajectories analysis. 
2. Do static analysis with all SDGs

Clean up a bit:

```{r}
rm(df_sdg, df_un, df_wb, maike_vars, other_vars)
# key files
# sdg raw data from UN SDG
# dat raw data from WorldBank
```

## Ordination

```{r}
pca1 <- df_dat |> 
    select(-country_name) |> 
    pivot_longer(cols = `Access to electricity (% of population)`:last_col(),
                 values_to = "value", names_to = "variable") |> 
    unite(col = "var", year, variable, sep = "_") |> 
    pivot_wider(names_from = var, values_from = value) |> 
    ungroup() |> 
    select(-country_code) |> 
    prcomp(center = TRUE, scale. = TRUE)
```

PCA
```{r}
plot(pca1)
```

```{r}
biplot(pca1)
```

```{r}
summary(pca1)
```

```{r}
pca2 <- df_un_reduced |> 
    unite(col = "var", series_code, year, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    ungroup() |> 
    select(-iso3) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    prcomp(center = TRUE, scale. = TRUE)
```

```{r}
biplot(pca2)
```
### Multiple factor analysis

```{r}
mfa1 <- df_un_reduced |> 
    unite(col = "var", series_code, year, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    ungroup() |> 
    select(-iso3) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 6), type = rep("s", 6)) # 18 years times 6 variables
```

```{r}
mfa1
```

```{r}
mfa1$eig[1:10,2] |> barplot(names = 1:10)

```
```{r}
mfa1
```

## To dos:

- one PCA per year, see how PCA1 change over time.

## Leftovers

Old code, most of the steps outlined are already implemented

```{r}
### World Bank dataset:
# vars is the list of variables with <30% missing values over time and space
# vars2 is the same list for UN dataset
# yrs <- sdg$time_period|> range(na.rm = TRUE) 
# 
# df_wb <- dat |>
#     filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
#     select(-x35) |> 
#     pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
#     mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
#     filter(year >= yrs[1], year <= yrs[2]) |> # use only years available in both datsets
#     rename(iso3 = country_code) |> select(-indicator_code) |> 
#     left_join(
#         select(un_countries, iso3 = `ISO-alpha3 Code`, country_name = `Country or Area`)
#     )
# 
# df_un <- sdg |> 
#     rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
#     filter(iso3 %in% countries_list, !is.na(value), indicator_name %in% vars2) |>
#     filter(goal %in% c(10, 14, 15)) |> 
#     group_by(indicator_name, year, iso3) |>
#     summarize(mean_val = mean(value, na.rm = TRUE)) |> 
#     ungroup() |> #group_by(indicator_name, year, iso3) |> 
#     # add_count() |> 
#     pivot_wider(names_from = year, values_from = mean_val) |> 
#     pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
#                  names_transform = list(year = as.numeric, value = as.numeric))|> 
#     left_join(
#         select(un_countries, iso3 = `ISO-alpha3 Code`, country_name = `Country or Area`)
#     )
#     
# df_sdg <- full_join(df_un, df_wb) |> 
#     rename(indicator = indicator_name)
# 
# df_sdg |> skimr::skim()
```

```{r}
# df_sdg |> 
#     ggplot(aes(year, value)) +
#     geom_line(aes(group = iso3), show.legend = FALSE) +
#     facet_wrap(~indicator, scales = "free_y")
```
remove countries with too few observations:

```{r}
# tot_vars <- df_sdg |> pull(indicator) |> unique() |> length()
# tot_yrs <- df_sdg |> pull(year) |> unique() |> length()
# tot_cnt <- df_sdg |> pull(iso3) |> unique() |> length()
# 
# df_sdg |> 
#     pivot_wider(names_from = indicator, values_from = value) |> 
#     pivot_longer(cols = 4:last_col(),names_to = "indicator", values_to = "value") |> 
#     mutate(nas = is.na(value)) |>
#     group_by(iso3, indicator) |> 
#     dplyr::summarize(missing = sum(nas)/ (tot_yrs)) |> 
#     summarize(suitable = missing > 0.3) |> 
#     ungroup() |> group_by(iso3) |> 
#     summarize(varsok = sum(suitable)) |> 
#     filter(varsok == tot_vars)
# 
# key_cntrs
```

Impute missing values:

```{r}
# df_sdg <- df_sdg |> 
#     filter(iso3 %in% key_cntrs) |> 
#     pivot_wider(names_from = indicator, values_from = value) |> 
#     pivot_longer(cols = 4:last_col(),names_to = "indicator", values_to = "value") |> 
#     group_by(indicator, iso3) |> 
#     mutate(value = na_interpolation(value, option = "stine"))
```

```{r}
# df_sdg |> 
#     ggplot(aes(year, value)) +
#     geom_line(aes(group = iso3), alpha = 0.5, show.legend = FALSE) +
#     facet_wrap(~indicator, scales = "free_y")
```
Check for correlations:

```{r}
# df_sdg |> 
#     pivot_wider(names_from = indicator, values_from = value) |> 
#     GGally::ggpairs(
#         columns = 4:21 ,
#     lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
#     diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
#     # upper = list(continuous = GGally::wrap("cor", size = 2))
#     ) + theme_light(base_size = 6)
```

```{r}
# df_sdg <- df_sdg |> 
#     ungroup() |> 
#     pivot_wider(names_from = indicator, values_from = value) |> 
#     mutate(
#         across(.cols = c("Cereal yield (kg per hectare)" ,"Forest area (sq. km)" ,
#             "Forest area certified under an independently verified certification scheme (thousands of hectares)" , "Total assistance for development, by donor countries (millions of current United States dollars)"  , "Total assistance for development, by recipient countries (millions of current United States dollars)"
#         ), .fns = log1p)
#     ) 
```

Paper title: the blind spots of sustainable development

- underscore the fact that a lot of variables are not available
- the kind of variables available are not appropriate in many cases
- sub-national conversation: countries interpret and collect data differently [carla and emmy work]
- what can be done to address the gaps? use of other datasets
- open questions for other SDGs dimensions...
